---
title: "Final Project"
author: "Annabelle Marsh"
date: "5/21/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(manipulate)
library(deSolve)
library(FME)
```

## Plot data

```{r}
# Input data, t in hours and P in absorbtion 
Ampicillin <- read_csv("Data.tsv")
t = Ampicillin$time
P0 = Ampicillin$control
ABSconversion = 8 * 10^8 #absorbance to CFUs/mL conversion - ratio taken from textbook
C=P0*ABSconversion #control data in CFUs/mL

P0.1 = Ampicillin$`0.1`
C0.1 = P0.1*ABSconversion #0.1 ug/mL concentration data in CFUs/mL

P0.5 = Ampicillin$`0.5`
C0.5 = P0.5*ABSconversion #0.5 ug/mL concentration data in CFUs/mL

P1 = Ampicillin$`1`
C1 = P1*ABSconversion #1 ug/mL concentration data in CFUs/mL

P3 = Ampicillin$`3`
C3 = P3*ABSconversion #3 ug/mL concentration data in CFUs/mL
```

```{r}
#plot of growth data from all concentrations
plot(t, C, xlab = "Time (minutes)", ylab = "Bacterial population (bacteria/mL)",
     ylim = c(0, 10e8), main = "Bacterial Growth in the Presense of Ampicillin")
points(t, C0.1, pch = 2)
points(t, C0.5, pch = 3)
points(t, C1, pch = 4)
points(t, C3, pch = 5)
legend("topleft", c("0 ug/mL", "0.1 ug/mL", "0.5 ug/mL", "1 ug/mL", "3 ug/mL"),
       pch = 1:5)
```

## Create the model and solve the control for carrying capacity and growth rate
```{r}
#solve the mathematical model
ABX = function(params, t, p0, D0){
  with(as.list(c(params)),{
    ABX.model= function(t,y,params){
      with(as.list(c(params,y)),{
        # model equations
        dpR = r*pR*(1- ((pR+pS)/K))              #bacterial growth, resistant to  ABX
        dpS = r*pS*(1- ((pR+pS)/K)) - alpha*A*pS #bacterial growth, susceptible
        dD = alpha*A*pS - delta*D              #dying phase
        list(c(dpR, dpS, dD))
      })}
    y0=c(pR=fracR*p0, pS=(1-fracR)*p0, D=D0)
    out=ode(y0,t,ABX.model,params)
    as.data.frame(out)
  })
}
#control
#no antibiotic and no antibacterial death, percent resistant irrelevant
ABX.error = function(unknown_params){
  with(as.list(unknown_params),{
    # solve ode system for a given set of parameters
    ABX.output = ABX(c(r = r, K = K, alpha = 0, delta = 0, A = 0, fracR=0), 
                     t, p0 = C[1], D0 = 0)  
    # return vector of squared residuals
    return(((ABX.output$pR+ABX.output$pS+ABX.output$D)-C)^2) #sum each population for total pop.
  })
}

#guesses from previously found params
unknown_params <- c(r = 0.02, K = 2.2e8)
fitControl = modFit(f=ABX.error, p = unknown_params, lower = c(0)) #minimize cost function
fitControl$par
r = fitControl$par[["r"]] # Save the parameters from the control fit
K = fitControl$par[["K"]]

#evaluate model at best fit parameters
modelSoln = ABX(c(r = r, K = K, alpha = 0, delta = 0, A = 0,
                  fracR = 0), t, p0 = C[1], D0 = 0)
#plot model
plot(t, C, ylab = "Population Size", xlab = "Time (minutes)", main = "Control", ylim = c(0,10e8))
lines(modelSoln$time, (modelSoln$pR+modelSoln$pS+modelSoln$D))
```
## Using the previously found carrying capacity and growth rate, solve all concentrations for the other parameters
```{r}
ABX.error = function(unknown_params){
  with(as.list(unknown_params),{
    #Same r and K from before, unknown params alpha, delta, and fracR (fraction resistant). A determined by concentration.
      # 0.1 ug/mL 
    ABX.outputC0.1 = ABX(c(r = r, K = K, alpha = alpha, 
                       delta = delta, A = 0.1, fracR = fracR), 
                     t = t, p0 = C0.1[1], D0 = 0)
      # 0.5 ug/mL 
    ABX.outputC0.5 = ABX(c(r = r, K = K, alpha = alpha, 
                       delta = delta, A = 0.5, fracR = fracR), 
                     t = t, p0 = C0.5[1], D0 = 0)
      # 1 ug/mL 
    ABX.outputC1 = ABX(c(r = r, K = K, alpha = alpha, 
                       delta = delta, A = 1, fracR = fracR), 
                     t = t, p0 = C1[1], D0 = 0)
      # 3 ug/mL 
    ABX.outputC3 = ABX(c(r = r, K = K, alpha = alpha, 
                       delta = delta, A = 3, fracR = fracR), 
                     t = t, p0 = C3[1], D0 = 0) 
    # return vector of squared residuals
    # sum models to get the total population, subtract the data, and then square
    return(
      # 0.1 ug/mL 
      (ABX.outputC0.1$pR+ABX.outputC0.1$pS+ABX.outputC0.1$D-C0.1)^2 +
      # 0.5 ug/mL
      (ABX.outputC0.5$pR+ABX.outputC0.5$pS+ABX.outputC0.5$D-C0.5)^2 + 
      # 1 ug/mL
      (ABX.outputC1$pR+ABX.outputC1$pS+ABX.outputC1$D-C1)^2 +
      # 3 ug/mL
      (ABX.outputC3$pR+ABX.outputC3$pS+ABX.outputC3$D-C3)^2) 
  })
}

unknown_params <- c(alpha = 0.02,
                       delta = 0.01, fracR = .05) # averaged results from previous fits to determine these params

# fit the model to all datasets besides control. 
# control is excluded because none of the parameters being fit impact the control due to A being 0.
# lower and upper bounds are because alpha and delta are rate parameters and fracR is a fraction.
fit = modFit(f=ABX.error, p = unknown_params, lower = c(0, 0, 0), upper = c(1,1,1)) 
```
## Fit all experimental datasets using previously determined r, K, alpha, and delta, but with unknown percent resistant (fracR)
### Fit 0.1ug/mL
```{r}
#0.1 micrograms/mL
ABX.error0.1 = function(unknown_params){
  with(as.list(unknown_params),{
    # solve ode system for a given set of parameters
    ABX.output = ABX(c(r = r, K = K, alpha = fit$par[["alpha"]], 
                       delta = fit$par[["delta"]], A = 0.1, fracR = fracR), 
                     t, p0 = C0.1[1], D0 = 0)  
    # return vector of squared residuals
    return(((ABX.output$pR+ABX.output$pS+ABX.output$D)-C0.1)^2) #sum each population for total pop.   
  })
}
#guesses from slider
unknown_params0.1 <- c(fracR = 0.2)
fit0.1 = modFit(f=ABX.error0.1, p = unknown_params0.1, lower = c(0), upper = c(1)) #minimize cost function
fit0.1$par
#just fit fracR, use previous fits for other parameters
modelSoln0.1 = ABX(params = c(r = r, K = K, alpha = fit$par[["alpha"]],
                      delta = fit$par[["delta"]], A = 0.1, fracR = fit0.1$par[["fracR"]]),
           t, p0 = C0.1[1], D0 = 0)
```
### Fit 0.5ug/mL
```{r}
#0.5 micrograms/mL 
ABX.error0.5 = function(unknown_params){
  with(as.list(unknown_params),{
    # solve ode system for a given set of parameters
    ABX.output = ABX(c(r = r, K = K, alpha = fit$par[["alpha"]], 
                       delta = fit$par[["delta"]], A = 0.5, fracR = fracR), 
                     t, p0 = C0.5[1], D0 = 0)  
    # return vector of squared residuals
    return(((ABX.output$pR+ABX.output$pS+ABX.output$D)-C0.5)^2) #sum each population for total pop.   
  })
}
#guesses from slider
unknown_params0.5 <- c(fracR = 0.1)
#just fit fracR, use previous fits for other parameters
fit0.5 = modFit(f=ABX.error0.5, p = unknown_params0.5, lower = c(0), upper = c(1)) #minimize cost function
fit0.5$par

modelSoln0.5 = ABX(params = c(r = r, K = K, alpha = fit$par[["alpha"]],
                      delta = fit$par[["delta"]], A = 0.5, fracR = fit0.5$par[["fracR"]]),
           t, p0 = C0.5[1], D0 = 0)
```
### Fit 1ug/mL
```{r}
#1 micrograms/mL, fix value (do not fit)
ABX.error1 = function(unknown_params){
  with(as.list(unknown_params),{
    # solve ode system for a given set of parameters
    ABX.output = ABX(c(r = r, K = K, alpha = fit$par[["alpha"]], 
                       delta = fit$par[["delta"]], A = 1, fracR = fracR), 
                     t, p0 = C1[1], D0 = 0)  
    # return vector of squared residuals
    return(((ABX.output$pR+ABX.output$pS+ABX.output$D)-C1)^2) #sum each population for total pop.   
  })
}
#guesses from slider
unknown_params1 <- c(fracR = 0.09)
fit1 = modFit(f=ABX.error1, p = unknown_params1, lower = c(0), upper = c(1)) #minimize cost function
fit1$par
#just fit fracR, use previous fits for other parameters
modelSoln1 = ABX(params = c(r = r, K = K, alpha = fit$par[["alpha"]],
                      delta = fit$par[["delta"]], A = 1, fracR = fit1$par[["fracR"]]),
           t, p0 = C1[1], D0 = 0)
```
### Fit 3ug/mL
```{r}
#3 micrograms/mL, fix value (do not fit)
ABX.error3 = function(unknown_params){
  with(as.list(unknown_params),{
    # solve ode system for a given set of parameters
    ABX.output = ABX(c(r = r, K = K, alpha = fit$par[["alpha"]], 
                       delta = fit$par[["delta"]], A = 3, fracR = fracR), 
                     t, p0 = C3[1], D0 = 0)  
    # return vector of squared residuals
    return(((ABX.output$pR+ABX.output$pS+ABX.output$D)-C3)^2) #sum each population for total pop.   
  })
}
#guesses from slider
unknown_params3 <- c(fracR = 0.04)
fit3 = modFit(f=ABX.error3, p = unknown_params3, lower = c(0), upper = c(1)) #minimize cost function
fit3$par
#just fit fracR, use previous fits for other parameters
modelSoln3 = ABX(params = c(r = r, K = K, alpha = fit$par[["alpha"]],
                      delta = fit$par[["delta"]], A = 3, fracR = fit3$par[["fracR"]]),
           t, p0 = C3[1], D0 = 0)
```
## Plot the final results
```{r}
#plot the results
#first, plot the data
plot(t, C, xlab = "Time (minutes)", ylab = "Bacterial population (bacteria/mL)",
     ylim = c(0, 10e8), main = "Model and Data")
points(t, C0.1, pch = 2)
points(t, C0.5, pch = 3)
points(t, C1, pch = 4)
points(t, C3, pch = 5)
#create the legend for the data
legend("topleft", c("0 ug/mL", "0.1 ug/mL", "0.5 ug/mL", "1 ug/mL", "3 ug/mL"),
       pch = 1:5)

#plot the models
lines(modelSoln$time, (modelSoln$pR+modelSoln$pS+modelSoln$D))
lines(modelSoln0.1$time, (modelSoln0.1$pR+modelSoln0.1$pS+modelSoln0.1$D), lty = 2)
lines(modelSoln0.5$time, (modelSoln0.5$pR+modelSoln0.5$pS+modelSoln0.5$D), lty = 3)
lines(modelSoln1$time, (modelSoln1$pR+modelSoln1$pS+modelSoln1$D), lty = 4)
lines(modelSoln3$time, (modelSoln3$pR+modelSoln3$pS+modelSoln3$D), lty = 5)
#create the legend for the models
legend("topleft", c("0 ug/mL", "0.1 ug/mL", "0.5 ug/mL", "1 ug/mL", "3 ug/mL"),
       pch = c(1,2,3,4,5), lty = c(1,2,3,4,5))
```

## Plot all parts of the model for 0.1 ug/mL 
```{r}
# plot the data
plot(t, C0.1, xlab = "Time (minutes)", ylab = "Bacterial population (bacteria/mL)",
     ylim = c(0, 10e8), main = "Model of 0.1 ug/mL Ampicillin")
# plot the total population model
lines(modelSoln0.1$time, (modelSoln0.1$pR+modelSoln0.1$pS+modelSoln0.1$D), col = 1)
# plot the resistant population model
lines(modelSoln0.1$time, modelSoln0.1$pR, col = 2)
# plot the susceptible popluation model
lines(modelSoln0.1$time, modelSoln0.1$pS, col = 3)
# plot the dying population model
lines(modelSoln0.1$time, modelSoln0.1$D, col = 4)
# create the legend
legend("topleft", c("Model", "Resistant", "Susceptible", "Dying", "Data"), col = c(1,2,3,4,1), lty = c(1,1,1,1,NA), pch = c(NA,NA,NA,NA,1))

```

## Plot all parts of the model for 0.5 ug/mL 
```{r}
# plot the data
plot(t, C0.5, xlab = "Time (minutes)", ylab = "Bacterial population (bacteria/mL)",
     ylim = c(0, 10e8), main = "Model of 0.5 ug/mL Ampicillin")
# plot the total population model
lines(modelSoln0.5$time, (modelSoln0.5$pR+modelSoln0.5$pS+modelSoln0.5$D), col = 1)
# plot the resistant population model
lines(modelSoln0.5$time, modelSoln0.5$pR, col = 2)
# plot the susceptible popluation model
lines(modelSoln0.5$time, modelSoln0.5$pS, col = 3)
# plot the dying population model
lines(modelSoln0.5$time, modelSoln0.5$D, col = 4)
# create the legend
legend("topleft", c("Model", "Resistant", "Susceptible", "Dying", "Data"), col = c(1,2,3,4,1), lty = c(1,1,1,1,NA), pch = c(NA,NA,NA,NA,1))

```

## Plot all parts of the model for 1 ug/mL 
```{r}
# plot the data
plot(t, C1, xlab = "Time (minutes)", ylab = "Bacterial population (bacteria/mL)",
     ylim = c(0, 10e8), main = "Model of 1 ug/mL Ampicillin")
# plot the total population model
lines(modelSoln1$time, (modelSoln1$pR+modelSoln1$pS+modelSoln1$D), col = 1)
# plot the resistant population model
lines(modelSoln1$time, modelSoln1$pR, col = 2)
# plot the susceptible popluation model
lines(modelSoln1$time, modelSoln1$pS, col = 3)
# plot the dying population model
lines(modelSoln1$time, modelSoln1$D, col = 4)
# create the legend
legend("topleft", c("Model", "Resistant", "Susceptible", "Dying", "Data"), col = c(1,2,3,4,1), lty = c(1,1,1,1,NA), pch = c(NA,NA,NA,NA,1))

```

## Plot all parts of the model for 3 ug/mL 
```{r}
# plot the data
plot(t, C3, xlab = "Time (minutes)", ylab = "Bacterial population (bacteria/mL)",
     ylim = c(0, 10e8), main = "Model of 3 ug/mL Ampicillin")
# plot the total population model
lines(modelSoln3$time, (modelSoln3$pR+modelSoln3$pS+modelSoln3$D), col = 1)
# plot the resistant population model
lines(modelSoln3$time, modelSoln3$pR, col = 2)
# plot the susceptible popluation model
lines(modelSoln3$time, modelSoln3$pS, col = 3)
# plot the dying population model
lines(modelSoln3$time, modelSoln3$D, col = 4)
# create the legend
legend("topleft", c("Model", "Resistant", "Susceptible", "Dying", "Data"), col = c(1,2,3,4,1), lty = c(1,1,1,1,NA), pch = c(NA,NA,NA,NA,1))

```
```{r}
fracRparams = c(fit3$par[["fracR"]], fit1$par[["fracR"]], fit0.5$par[["fracR"]], fit0.1$par[["fracR"]])
concentration = c(3,1,0.5,0.1)
plot(concentration, fracRparams * 100,  xlab = "Ambicillin Concentration ug/mL", ylab = "Percent Resistant")
```
